---
title: "Danish Kings"
author: "Clara Laursen & Liva Rask"
date: "2026-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator: , 

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("Data/kings.csv")

kings2 <- read_csv("Data/kings.csv")

kings3 <- read.csv2("Data/kings.csv")

kings4 <- read_csv2("Data/kings.csv")
```
```{r}
class(kings1)
class(kings2)
class(kings3)
class(kings4)
```

Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object: `read_csv()` is the tidyverse function.
2. What is the result of running `class()` on the `kings` object created with a tidyverse function: Running `class()` shows that `read.csv()` creates a data frame, while `read_csv()` creates a tibble.
3. How many columns does the object have when created with these different functions?: The `kings` object created with `read_csv()` has 11 columns.
4. Show the dataset so that we can see how R interprets each column: 

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv("Data/kings.csv")

class(kings)


head(kings)
```
## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`

Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.



```{r warning=FALSE}
kings <- kings %>%
  mutate(
    Start_year = as.numeric(Start_year),
    End_year = as.numeric(End_year),
    duration = End_year - Start_year
  )

head(kings)
```

I used `mutate()` to create a new column called `duration` by subtracting `Start_year` from `End_year`.

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`

```{r}
mean(kings$duration, na.rm = TRUE)

kings %>%
  summarize(avg_duration = mean(duration, na.rm = TRUE))
```
I selected the `duration` column and calculated the average with `mean(..., na.rm = TRUE)`. I also used `summarize()` to calculate the same result on the tibble.

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
avg_duration <- mean(kings$duration, na.rm = TRUE)

kings %>%
  filter(duration > avg_duration) %>%
  select(Name, duration)

kings %>%
  filter(duration > avg_duration) %>%
  count()
```
25 kings had a longer-than-average reign. I identified them by filtering the `duration` column for values above the average.

## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
kings %>%
  arrange(desc(duration)) %>%
  slice(1:3) %>%
  mutate(Days = duration * 365) %>%
  select(Name, duration, Days)
```
The three longest-ruling monarchs were Christian 4., Margrete 2., and Erik 7. af Pommern. Their estimated reigns lasted 21,900, 18,615, and 15,695 days. 

## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
kings <- kings %>%
  mutate(midyear = Start_year + (End_year - Start_year) / 2)

ggplot(kings, aes(x = midyear, y = duration)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Duration of reign among Danish monarchs over time",
    x = "Midpoint of reign",
    y = "Duration of reign (years)"
  ) +
  theme_bw()
```
The plot suggests a slight long-term increase in reign duration over time, although there is a lot of variation. Some monarchs in later periods had especially long reigns.
