---
title: "Sentiment Analysis of Game of Thrones by Clara Laursen & Liva Rask."
date: 'updated `r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(here)

# For text mining:

library(pdftools)
library(tidytext)
library(textdata) 
library(ggwordcloud)
```
## Research question

What does the distribution of AFINN sentiment scores suggest about the emotional tone of Game of Thrones?

## Method

The text was extracted from the PDF, tokenized into words, cleaned by removing stop words and numbers, and analysed with the AFINN and NRC sentiment lexicons.

## Data
```{r get-document}
got_path <- here("data","got.pdf")
got_text <- pdf_text(got_path)
```

## Cleaning and structuring the text

The extracted PDF text was split into lines, unnested into rows, and trimmed to remove extra whitespace before tokenization.

```{r split-lines}
got_df <- data.frame(got_text) %>%
  mutate(text_full = str_split(got_text, pattern = '\n')) %>%
  unnest(text_full) %>%
  mutate(text_full = str_trim(text_full))
```
Each line from the PDF was stored as a separate row.

### Preparing words for analysis

The text was tokenized into individual words using `unnest_tokens()`.

```{r tokenize}
got_tokens <- got_df %>%
  unnest_tokens(word, text_full)
```

Counting the words:
```{r count-words}
got_wc <- got_tokens %>%
  count(word) %>%
  arrange(-n)
```

### Removing stop words:
Common stop words were removed using `anti_join(stop_words)`. The remaining words were then counted again to identify the most frequent meaningful terms in the text.

 `tidyr::anti_join()`:
```{r stopwords}
got_stop <- got_tokens %>%
  anti_join(stop_words)
```

```{r count-words2}
got_swc <- got_stop %>%
  count(word) %>%
  arrange(-n)
```

## Removing numeric tokens
```{r skip-numbers}
got_no_numeric <- got_stop %>%
  filter(is.na(as.numeric(word)))
```
Numeric tokens were removed so that the analysis focused on words rather than page numbers or other non-text elements.

##### Preparing data for the word cloud
This word cloud visualises the most frequent meaningful words in *Game of Thrones*.
```{r wordcloud-prep}
length(unique(got_no_numeric$word))

got_top100 <- got_no_numeric %>%
  count(word) %>%
  arrange(-n) %>%
  head(100)

got_top100
```

```{r wordcloud}
got_cloud <- ggplot(data = got_top100, aes(label = word)) +
  geom_text_wordcloud() +
  theme_minimal()

got_cloud
```

This version highlights the most frequent words more clearly by scaling word size and colour according to frequency: 
```{r wordcloud-pro}
ggplot(data = got_top100, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "star") +
  scale_size_area(max_size = 12) +
  scale_color_gradientn(colors = c("darkgreen", "blue", "red")) +
  theme_minimal()
```
## Sentiment analysis
To analyse the emotional vocabulary of *Game of Thrones*, I used the AFINN and NRC lexicons. AFINN assigns words scores from negative to positive, while NRC groups words into emotion categories such as fear, trust, joy, and sadness.

```{r bind-afinn}
got_afinn <- got_stop %>%
  inner_join(get_sentiments("afinn"))
got_afinn
```

The matched words were counted by AFINN score to show the overall distribution of negative and positive terms in the text.
```{r count-afinn}
got_afinn_hist <- got_afinn %>% 
  count(value)

ggplot(data = got_afinn_hist, aes(x = value, y = n)) +
  geom_col(aes(fill = value)) +
  theme_bw()
```

Investigate some of the words in a bit more depth:
```{r afinn-2}

got_afinn2 <- got_afinn %>% 
  filter(value == 2)
```

```{r afinn-2-more}
unique(got_afinn2$word)

got_afinn2_n <- got_afinn2 %>% 
  count(word, sort = TRUE) %>% 
  mutate(word = fct_reorder(factor(word), n))


ggplot(data = got_afinn2_n, aes(x = word, y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw()
```
### Overall AFINN summary
To summarise the overall sentiment tendency in the text, I calculated the mean and median AFINN scores of the matched words.

The mean and median indicate *slightly* positive overall sentiments based on the AFINN lexicon. 

```{r summarize-afinn}
got_summary <- got_afinn %>% 
  summarize(
    mean_score = mean(value),
    median_score = median(value)
  )
got_summary
```

## NRC sentiment analysis
The NRC lexicon was used to group words in *Game of Thrones* into emotion categories such as fear, trust, joy, sadness, and positive/negative sentiment.

```{r bind-bing}
got_nrc <- got_stop %>% 
  inner_join(get_sentiments("nrc"))
```

The excluded words were counted to identify which frequent terms were missing from the NRC lexicon.
```{r check-exclusions}
got_exclude <- got_stop %>%
  anti_join(get_sentiments("nrc"))

got_exclude_n <- got_exclude %>%
  count(word, sort = TRUE)

head(got_exclude_n)
```

Now find some counts: 
```{r count-bing}
got_nrc_n <- got_nrc %>% 
  count(sentiment, sort = TRUE)


ggplot(data = got_nrc_n, aes(x = sentiment, y = n)) +
  geom_col(aes(fill = sentiment))+
  theme_bw()
```

The most frequent words within each NRC sentiment category were plotted to show which terms drive the emotional categories.
```{r count-nrc}
got_nrc_n5 <- got_nrc %>% 
  count(word,sentiment, sort = TRUE) %>% 
  group_by(sentiment) %>% 
  top_n(5) %>% 
  ungroup()

got_nrc_gg <- ggplot(data = got_nrc_n5, aes(x = reorder(word,n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, ncol = 2, scales = "free") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Word", y = "count")

got_nrc_gg

ggsave(plot = got_nrc_gg, 
       here("figures","week13_got_nrc_sentiment.png"), 
       height = 8, 
       width = 5)

```
Some words in the NRC categories are context-dependent, which shows that lexicon-based sentiment analysis should be interpreted critically.
